%\documentclass{bioinfo}
%\copyrightyear{2019} \pubyear{2019}

%\access{Advance Access Publication Date: Day Month Year}
%\appnotes{Applications Note}

\documentclass[11pt]{article}

\usepackage{amsmath, amsfonts, amsthm}
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]

\usepackage[margin=1.25in]{geometry}
\usepackage{setspace}
\onehalfspacing
\usepackage{subfigure}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{authblk}

\setcounter{tocdepth}{2}

\begin{document}
%\firstpage{1}

%\subtitle{Genetic and population analysis}

% \title[Supplement to Unified GWAS power analysis]{Supplement to ``U-PASS: unified power analysis and forensics for qualitative traits in GWAS''}
\title{Supplement to ``U-PASS: unified power analysis and forensics for qualitative traits in gentic association studies''}
\author[1]{Zheng Gao\thanks{Corresponding author: gaozheng@umich.edu}}
\author[1]{Jonathan Terhorst}
\author[2]{Cristopher Van Hout}
\author[1]{Stilian Stoev}
\affil[1]{Department of Statistics, University of Michigan, Ann Arbor, MI 48109, USA}
\affil[2]{Regeneron Genetics Center, Tarrytown, NY 10591 USA.}
%\author[Gao \textit{et~al}.]{Zheng Gao\,$^{\text{\sfb 1,}*}$, Jonathan Terhorst\,$^{\text{\sfb 1}}$, Cris Van Hout\,$^{\text{\sfb 2}}$, and Stilian Stoev\,$^{\text{\sfb 1}}$}
%\address{$^{\text{\sf 1}}$Department of Statistics, University of Michigan, Ann Arbor, MI 48109, USA and \\
%$^{\text{\sf 2}}$Regeneron Genetics Center, Basking Ridge, NJ 07920, USA.}

%\corresp{$^\ast$To whom correspondence should be addressed.}

%\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

%\editor{Associate Editor: XXXXXXX}

\maketitle

\vspace{-20pt}
\begin{abstract}
   We provide a brief overview of power analysis for genetic association studies in Section \ref{sec:power-analysis-overview}.
   The relationship between two approaches to specifying the alternative hypothesis in power calculations for association studies is clarified in Section \ref{sec:disease-models}.
   We develop a unified analysis of statistical power for common association tests in Section \ref{sec:asymptotic-equivalence}, and provide a simple rule for finite-sample correction of the asymptotic approximations in Section \ref{sec:finite-sample}.
   The performance of this correction is examined through numerical experiments in Section \ref{sec:numerical}.
   Proof of the main result is found in Section \ref{sec:proof}.
\end{abstract}

\tableofcontents

\section{Overview of power analysis in genetic association studies}
\label{sec:power-analysis-overview}
\input{0.A.power-analysis-overview.tex}

\section{Two approaches to specifying the alternative hypothesis}
\label{sec:disease-models}
\input{0.B.disease-models.tex}

\section{Use case illustrations}
\label{sec:user-guide}
\input{0.C.use-case-illustrations.tex}

\section{Unified asymptotic analysis of power}
\label{sec:asymptotic-equivalence}
\input{0.D.asymptotic-equivalence.tex}

\section{Finite-sample corrections}
\label{sec:finite-sample}
\input{0.E.finite-sample.tex}

%\vspace{*{-5pt}
\section{Proof of Theorem \ref{thm:1}}
\label{sec:proof}
\input{0.F.proofs.tex}

%\vspace{{10pt}
\section{Numerical illustrations}
\label{sec:numerical}
\input{0.G.numerical.tex}

% \section{An errata for Johnson et~al.\ (2017)}
% \label{sec:errata}
% \input{0.H.errata.tex}

% \vspace*{-10pt}
% \section*{Acknowledgements}
% 
% We thank so and so for their invaluable comments and suggestions on the early versions of the software and this paper.

%\vspace{*{-12pt}
\section*{Funding}

This work is partially supported by NSF Grant DMS-1830293, Algorithms for Threat Detection.


%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

%\vspace{*{-12pt}
\begin{thebibliography}{}

\normalsize{
\bibitem[Choi {\it et~al}., 2017]{Choi17}
Choi, L., Blume, J. D., \& Dupont, W. D. (2015). 
Elucidating the foundations of statistical inference with 2 x 2 tables. 
{\it PloS one}, {\bf 10(4)}, e0121263.

\bibitem[Derkach {\it et~al}., 2017]{Derkach17}
Derkach, A., Zhang, H., \& Chatterjee, N. (2017). 
Power Analysis for Genetic Association Test (PAGEANT) provides insights to challenges for rare variant association studies. 
{\it Bioinformatics}, {\bf 34(9)}, 1506-1513.

\bibitem[Dominguez-Cruz {\it et~al}., 2018]{Dominguez-Cruz18}
Dom\'inguez-Cruz, M. G., de Lourdes Mu\~noz, M., Totomoch-Serra, A., Garc\'ia-Escalante, M. G., Burgue\~no, J., Valadez-Gonz\'alez, N., ... \& D\'iaz-Badillo, \'A. (2018). 
Pilot genome-wide association study identifying novel risk loci for type 2 diabetes in a Maya population. 
{\it Gene}, {\bf 677}, 324-331.

\bibitem[Ferguson, 2017]{Ferguson17}
Ferguson, T. S. (2017). {\it A course in large sample theory}. Routledge.

\bibitem[Gonz\'{a}lez {\it et~al}., 2008]{Gonzalez08}
Gonz\'{a}lez, J. R., Carrasco, J. L., Dudbridge, F., Armengol, L., Estivill, X., \& Moreno, V. (2008). 
Maximizing association statistics over genetic models. 
{\it Genetic Epidemiology}, {\bf 32(3)}, 246-254.

\bibitem[Hunter, 2002]{Hunter02}
Hunter, D. (2002) Topic 25: Asymptotic Power of Pearson’s Chi-Square Test in 
{\it Statistics 597A: Asymptotic Tools (Fall 2002), PSU Lecture notes,} http://personal.psu.edu/drh20/asymp/fall2002/lectures/ln11.pdf

\bibitem[Johnson {\it et~al}., 2017]{Johnson17}
Johnson, J. L., \& Abecasis, G. R. (2017). 
GAS Power Calculator: web-based power calculator for genetic association studies. 
{\it bioRxiv}, 164343.

\bibitem[Lehmann, 2004]{Lehmann04}
Lehmann, E. L. (2004). {\it Elements of large-sample theory}. Springer Science \& Business Media.

\bibitem[Li {\it et~al}., 2008]{Li08}
Li, Q., Zheng, G., Li, Z., \& Yu, K. (2008). 
Efficient approximation of P‐value of the maximum of correlated tests, with applications to genome‐wide association studies. 
{\it Annals of human genetics}, {\bf 72(3)}, 397-406.

\bibitem[MacArthur {\it et~al}., 2016]{MacArthur16}
MacArthur, J., Bowler, E., Cerezo, M., Gil, L., Hall, P., Hastings, E., ... \& Pendlington, Z. M. (2016). 
The new NHGRI-EBI Catalog of published genome-wide association studies (GWAS Catalog). 
{\it Nucleic acids research}, {\bf 45(D1)}, D896-D901.

\bibitem[Menashe {\it et~al}., 2008]{Menashe08}
Menashe, I., Rosenberg, P. S., \& Chen, B. E. (2008). 
PGA: power calculator for case-control genetic association analyses. 
{\it BMC genetics}, {\bf 9(1)}, 36.

\bibitem[Purcell {\it et~al}., 2003]{Purcell03}
Purcell, S., Cherny, S. S., \& Sham, P. C. (2003). 
Genetic Power Calculator: design of linkage and association genetic mapping studies of complex traits. 
{\it Bioinformatics}, {\bf 19(1)}, 149-150.

\bibitem[Ripamonti {\it et~al}., 2017]{Ripamonti17}
Ripamonti, E., Lloyd, C., \& Quatto, P. (2017). 
Contemporary Frequentist Views of the $2\times2$ Binomial Trial. 
{\it Statistical Science}, {\bf 32(4)}, 600-615.

\bibitem[Sham, 1998]{Sham98}
Sham, P. C. (1998). {\it Statistics in human genetics}. Wiley.

\bibitem[Skol {\it et~al}., 2006]{Skol06}
Skol, A. D., Scott, L. J., Abecasis, G. R., \& Boehnke, M. (2006). 
Joint analysis is more efficient than replication-based analysis for two-stage genome-wide association studies. 
{\it Nature genetics}, {\bf 38(2)}, 209.

\bibitem[Smyth {\it et~al}., 2017]{Smyth17}
Smyth, G., Hu, Y., Dunn, P., Phipson, B., \& Chen, Y. (2017)
statmod: Statistical Modeling.
R package version 1.4.30.
%{\it https://cran.r-project.org/web/packages/statmod}.

\bibitem[Wang {\it et~al}., 2014]{Wang14}
Wang, G. T., Li, B., Lyn Santos-Cortez, R. P., Peng, B., \& Leal, S. M. (2014).
Power analysis and sample size estimation for sequence-based association studies. 
{\it Bioinformatics}, {\bf 30(16)}, 2377-2378.
}

\end{thebibliography}
\end{document}
